---
title: "Motion tracking analysis"
author: "Amalie Lysgaard Andersen"
date: "4/4/2019"
output: html_document
---

```{r setup, include=FALSE}
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/4. Aarhus Universitet/4. Semester/3. Models for perception and action/Tasks_ModelsMindPerception")

pacman::p_load(readr, tidyverse, reshape, ggplot2)
```

### Motion tracking and analysis (contd.) ###
In your study group, try to do the following analysis steps with a single trial of the hand priming data in R. Upload your commented code to Blackboard (or share it with me in some other way). If you get stuck anywhere make a comment about it – just see how far you get! We’ll discuss it together during the next class.
Deadline for submission: Tuesday, April 9th.

Note: Use the data 21, not any of the others I had previously uploaded. The data folder contains all 20 trials of that one participant. The data are already filtered.

Matlab treats the top as 0.0, which means that the data is then inverted (U shapes instead of n shapes) --> cordula already did this

```{r data}
#loading
data <- list.files(pattern = "*.txt") %>% 
  lapply(read.delim, stringsAsFactors=F, sep=",", header=F) %>% 
   bind_rows 

#renaming
data <- rename(data, c(V1 = "sample", V2 = "condition", V3 = "pos_x", V4 = "pos_y"))

single_15 <- list.files(pattern = "21_15.txt") %>% 
  lapply(read.delim, stringsAsFactors=F, sep=",", header=F)
```


#1. Remind yourselves what we did in the study and what we expected and why.
Mouse tracking: opstacle course. Is there a pattern in how high we go in the obstacles when changing the height pattern?

- Mouse-tracking: like a histogram, taking a course without touching the pillars
- Intention with experiment: 
	• Gauge the repetition effect
	• If deviation from the pattern (difference in height) --> longer RT
	• Remapping effect
	• If many tall pillars, and one low in the middle --> you would probably go a lot higher on the low one as you're used to the taller ones
	  	○ Cordula found this effect!
	  	○ Repetition --> it's cost-effective to just keep the previous pattern
	• If many low pillers, and one tall one in the middle --> you overestimate the sufficient movement
	  	○ The tall pillar is an obstacle 

The data is more mixed in this effect though

# 2. Load one trial of your choice into R. The four columns refer to: sample, condition, x and y.
```{r}
single <- read.delim("21_15.txt", sep = ",", header =F)

single <- rename(single, c(V1 = "sample", V2 = "condition", V3 = "pos_x", V4 = "pos_y"))
```


# 3. Plot x against y.
```{r}
ggplot(single, aes(x=pos_x, y=pos_y)) +
  geom_point(size = 0.2,color = rainbow(2268))
```


4. Map the y data to a more intuitive screen output with 0/0 in the lower left corner and where
upwards movements are increasingly positive (i.e. so that it looks like the movements we made
on the screen).
```{r}
single$pos_y_scale <- single$pos_y - min(single$pos_y)
single$pos_x_scale <- single$pos_x - min(single$pos_x)


ggplot(single, aes(x=pos_x_scale, y=pos_y_scale)) +
  geom_point(size = 0.2,color = rainbow(2268))
```

#5. Calculate the distance travelled (for x and y combined).
calculate euclidian distance....
(note: there's two dimensions to take into account here --> it's not just one linear line)
```{r}
correct_manualdist <- sum(sqrt(diff(single$pos_x_scale)^2 + diff(single$pos_y_scale)^2))
correct_manualdist

# 5523.131 pixels: the distance from start to end of trial --> this is the actual distance the mouse is travelling
```


#6. Add a time vector. I forgot to measure time so I had to estimate it post-hoc (fine for this exercise, not great if it was a real experiment). We can assume that each sample takes .0025 s.
```{r}
single$time <- (single$sample - 1) * 0.0025 
```


#7. Calculate velocity and acceleration for y, i.e. vertical movement. Remember: Velocity = distance difference / time difference. Acceleration = velocity difference / time difference.
```{r}
single$dist_diff <- single$pos_y - lag(single$pos_y, n = 1)
single$velocity <- (single$pos_y - lag(single$pos_y, n = 1)) / 0.0025

single$velo_diff <- single$velocity - lag(single$velocity, n = 1)
single$acceleration <- (single$velocity - lag(single$velocity, n = 1))/0.0025
```


#8. Play around with some filter to smoothen velocity and acceleration (most common: Butterworth filter).
```{r}
# first plotting

# acceleration
ggplot(single, aes(x=time, y=acceleration)) +
  geom_point(size = 0.2,color = rainbow(2268))

# velocity
ggplot(single, aes(x=time, y=velocity)) +
  geom_point(size = 0.2,color = rainbow(2268))

```



#9. Using zero crossings in the acceleration (EDIT: 0-CROSSINGS IN VELOCITY), calculate velocity extreme values (maxima and minima). Mark those extreme values on top of the x/y plot you made earlier.
A messy acceleration curve --> one point at the most extreme value wouldn't be that informative
```{r}



```


#10. Using e.g. a range of x positions, specify a segment during which the critical movement happens, i.e. the 6th movement is what we would want to compare in this study. Mark the peak in that segment in a different color.                                               
```{r}

```


#Bonus: If you want, analyse and plot all 20 trials separately for each condition.
